```{r}
library(tidyverse)
library(data.table)
library(here)
library(ggrepel)
library(patchwork)

theme_set(theme_bw())
```

```{r}
K562.bulk <- fread(here("data/perturb/pairs/K562/bulk_gaussian_trans_dat.csv"))
K562.perturb <- fread(here("data/perturb/pairs/K562/bulk_pertubration_pairs.csv"))
K562.qtl <- fread(here("data/perturb/pairs/K562/bulk_eQTL_pairs.csv"))
```

```{r}
Jurkat.bulk <- fread(here("data/perturb/pairs/Jurkat/bulk_gaussian_trans_dat.csv"))
Jurkat.perturb <- fread(here("data/perturb/pairs/Jurkat/bulk_pertubration_pairs.csv"))
Jurkat.qtl <- fread(here("data/perturb/pairs/Jurkat/bulk_eQTL_pairs.csv"))
```

Perturbation efficiencies
```{r}
Jurkat.perturb %>% filter(perturbation == effect) %>% pull(avg_log2FC) %>% summary
K562.perturb %>% filter(perturbation == effect) %>% pull(avg_log2FC) %>% summary
```

```{r}
K562.efficiency <- K562.perturb %>% filter(perturbation == effect) %>% mutate(KD_eff = 2^avg_log2FC)
Jurkat.efficiency <- Jurkat.perturb %>% filter(perturbation == effect) %>% mutate(KD_eff = 2^avg_log2FC)

K_eff <- ggplot(K562.efficiency, aes(x=KD_eff)) +
  geom_density() +
  ggtitle("K562 Knockdown efficiency") +
  xlab("KD efficiency (expression % from baseline)") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red")  # efficiency 70%
J_eff <- ggplot(Jurkat.efficiency, aes(x=KD_eff)) +
  geom_density() +
  ggtitle("Jurkat Knockdown efficiency") +
  xlab("KD efficiency (expression % from baseline)") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red")  # efficiency 70%

p <- K_eff + J_eff + plot_annotation(tag_levels = 'A')

ggsave(here("plots/exploratory/perturbation_efficiency.png"), plot = p, height = 4, width = 8)
```

```{r}
K562.efficiency$KD_eff %>% summary
```

```{r}
Jurkat.efficiency$KD_eff %>% summary
```

Jurkat has better efficiencies so we'll focus on it first. No need to filter.

Select only SNPs with strong negative cis effect
Select the 20 SNPs with the most negative cis effect

```{r}
top20 <- Jurkat.qtl %>%
  group_by(SNP) %>%
  slice_min(Beta.perturb, with_ties = FALSE) %>%  # keep most negative per SNP
  ungroup() %>%
  slice_min(Beta.perturb, n = 20) %>%
  pull(SNP)

strong.qtl <- Jurkat.qtl %>% filter(SNP %in% top20)
```


Those are QTL pairs with strong cis-eQTL which is strongly negative and would have similar effect to perturbation
```{r}
strong <- merge(strong.qtl %>% select(Beta.perturb, Beta.effect, perturbation, effect), Jurkat.perturb, by=c("perturbation", "effect"))
```

```{r}
# select only the row with the minimum Beta.perturb per perturbation–effect pair
strong <- strong %>%
  group_by(perturbation, effect) %>%
  slice_min(Beta.perturb, with_ties = FALSE) %>%
  ungroup()

# 1. Define quadrants
strong <- strong %>%
  mutate(quadrant = case_when(
    Beta.effect > 0 & avg_log2FC > 0 ~ "Q1",
    Beta.effect < 0 & avg_log2FC > 0 ~ "Q2",
    Beta.effect < 0 & avg_log2FC < 0 ~ "Q3",
    Beta.effect > 0 & avg_log2FC < 0 ~ "Q4",
    TRUE ~ "Other"
  ))

# 2. Calculate percentages
quad_counts <- strong %>%
  filter(quadrant != "Other") %>%
  count(quadrant) %>%
  mutate(percentage = round(100 * n / sum(n), 1))

# 3. Set quadrant label positions
quad_labels <- tibble(
  quadrant = c("Q1", "Q2", "Q3", "Q4"),
  x = c(0.3, -0.3, -0.3, 0.3),    # approximate x positions
  y = c(1, 1, -1, -1)     # approximate y positions
) %>%
  left_join(quad_counts, by = "quadrant") %>%
  mutate(label = paste0(percentage, "%"))



# 4. Plot with annotations
ggplot(strong, aes(x = Beta.effect, y = avg_log2FC)) +
  geom_point(size = 0.1) +
  geom_hline(yintercept = 0, color = "red") +
  geom_vline(xintercept = 0, color = "red") +
  ylab("log2FC") +
  xlab("trans-eQTL effect size") + 
  ggtitle("Relationship between trans-eQTL effect size and perturb-seq") +
  geom_text(data = quad_labels, aes(x = x, y = y, label = label), 
            color = "blue", size = 4, fontface = "bold") +
  theme_bw()



strong.filter_extreme <- strong %>% filter(avg_log2FC > -15)
ggplot(strong.filter_extreme, aes(x=Beta.effect, y=avg_log2FC)) +
  geom_point(size=0.1) +
  geom_hline(yintercept = 0,color ="red") +
  geom_vline(xintercept = 0,color ="red") +
  ylab("log2FC") +
  xlab("trans-eQTL effect size") + 
  ggtitle("Filter the -15 to see the rest") +
  geom_text(data = quad_labels, aes(x = x, y = y, label = label), 
            color = "blue", size = 4, fontface = "bold") +
  theme_bw()


```
Signed regulator-trans-eQTL correlation vs trans-eQTL effect size (like figure 3C in Pritchard's paper)

```{r}
strong <- strong %>% mutate(strong)
```



```{r}
top20 <- Jurkat.qtl %>%
  group_by(SNP) %>%
  slice_min(Beta.perturb, with_ties = FALSE) %>%  # keep most negative per SNP
  ungroup() %>%
  slice_min(Beta.perturb, n = 20) %>%
  pull(SNP)
strong.qtl <- Jurkat.qtl %>% filter(SNP %in% top20)

strong <- merge(strong.qtl %>% select(Beta.perturb, Beta.effect, perturbation, effect), Jurkat.perturb, by=c("perturbation", "effect"))
```

```{r}

Jurkat.qtl %>% group_by(perturbation) %>% summarise(uniSNP = length(unique(SNP))) %>% arrange(uniSNP)
strong$perturbation %>% table
strong %>% group_by(perturbation) %>% summarise(avglfc = mean(avg_log2FC)) %>% arrange(avglfc)
```

